% ===========================================================================
\chapter{Literature Review}
% ===========================================================================

The title of this thesis --- \textit{Intelligent Security Posture
Aware Service Function Chain Placement Using Deep Reinforcement
Learning} --- commits to four coupled claims: (i) that \textbf{SFC
placement} is a technically meaningful research problem; (ii) that an
\textbf{intelligent, learning-based} placement policy offers
advantages over classical heuristic and optimization approaches; (iii)
that \textbf{security posture} is a
first-class placement concern that must be modelled explicitly; and
(iv) that \textbf{deep reinforcement learning (DRL)}, appropriately
extended with hard-constraint masking and topology-aware policy
representations, is the right class of intelligent method to deliver
all of the above simultaneously.

This chapter validates each of these four claims against the recent
literature. Section~\ref{sec:lr_sfc_placement_problem} establishes
that SFC placement is an active, unresolved research problem whose
multi-dimensional constraints justify sustained study.
Section~\ref{sec:lr_security_aware} surveys security-aware placement
work. Section~\ref{sec:lr_drl} reviews DRL-based placement and
shows both that DRL is the emerging technique of choice and that no
prior DRL work is simultaneously security-aware, hard-constrained, and
topology-aware. Section~\ref{sec:lr_classical} surveys classical
optimization and heuristic approaches to motivate the
\textit{intelligent, learning-based} stance taken by the thesis.
Finally, Section~\ref{sec:lr_synthesis} presents a comparative
summary, mapping each of the four title components to concrete gaps
in the literature and showing how the proposed framework is the
first, to our knowledge, to close all four simultaneously.

\section{SFC Placement: A Multi-Constraint Online Problem}
\label{sec:lr_sfc_placement_problem}

The first claim implicit in the title is that \textit{SFC placement}
continues to warrant dedicated research. The breadth of activity
surveyed in this chapter confirms it: recent work attacks SFC
placement through heuristics, meta-heuristics, mathematical
programming, learning-augmented optimization, and deep reinforcement
learning --- each motivated by a different combination of objectives
(delay, cost, energy, reliability, security) and deployment contexts
(cloud, edge, fog, multi-domain, multi-vendor, IoT, air traffic
networks). Saidi~et~al. prove that VNF placement and chaining remains
NP-complete even under restrictive assumptions on infrastructure and
chain size, motivating the continued search for efficient online
methods~\cite{saidi_constrained_2024}; Han, Meng, and Fan show that
even basic choices such as ``cloud versus edge'' placement have
non-trivial consequences for substrate
utilization~\cite{han_sfc_2024}; Liang~et~al. demonstrate that
seemingly lower-level concerns such as the \emph{position} of packet
loss within a chain can dominate end-to-end
cost~\cite{liang_position-aware_2024}. The problem therefore is not
merely open, it is multi-dimensional, dynamic, and rich with
sub-problems that continue to generate new formulations.

\section{Security-Aware SFC Placement}
\label{sec:lr_security_aware}

The second title claim --- that placement must be \textit{security
posture aware} --- begins with the observation that security is now a
first-class placement objective in virtualized networks. The surveyed
works in this area differ along two primary axes: (i) how the security
requirement is modelled, and (ii) what is optimized (cost, delay, revenue,
protection).

\subsection{Static-Level Security and Delay Optimization}

An influential strand of work treats security as a static property of
each substrate node against which a service-specified minimum level
must be enforced. Yang~et~al. develop an air-traffic-information
network placement method, ESSFCD-DO, in which VNF aggregation
(VNFAA-SRCO), topology-aware VNF deployment (VNFD-TA), and $k$-shortest
path routing are composed to minimize end-to-end delay while
respecting security and resource budgets; the authors report
improvements in both delay and long-term revenue-to-cost
ratio~\cite{yang_efficient_2024}. Sun~et~al. study the same family of
objectives for edge-cloud task scheduling of DNN-based IoT services,
combining a security-related risk model with time-guaranteed domain
construction and security-aware task ranking, and demonstrating gains
over several delay- or security-only
baselines~\cite{sun_security-aware_2023}.

While effective, both approaches rely on heuristic methods that solve
each request in isolation and cannot adapt to evolving traffic or
substrate conditions.

\subsection{Security-Aware Scheduling and Cost Optimization}

Dubba and Killi formulate the dynamic SFC scheduling problem with
simultaneous security, cost, and deadline constraints and prove it
NP-hard. They propose two particle-swarm heuristics --- PSO-SFCS and a
group-learning variant GLPSO-SFCS --- over a three-level security
model that distinguishes the security posture of both VNFs and
physical machines, reporting sizable reductions in security-violation,
deadline-violation, and SLA-violation
ratios~\cite{dubba_security-aware_2024}. Wang~et~al. consider
security-aware SFC embedding with \textit{asymmetric dedicated
protection}: each service function has multiple versions with
different security guarantees and costs, and their APPF-DSI augmenting-path
algorithm jointly chooses the version, builds a primary-first
disjoint backup chain, and embeds both, improving cost over benchmarks
extended from the state of the
art~\cite{wang_security-aware_2024}. Wang~et~al. extend this line of
work to multi-vendor edge networks, identifying the ``redundancy
security accumulation'' phenomenon caused by directly summing the
security levels of chained functions, and propose the
security-cost-balance SCB-SD heuristic, which is shown to be a
logarithmic-approximation of the
optimum~\cite{wang_towards_2025}. Both Wang contributions integrate
security into the \emph{composition} of the chain itself, not just the
\emph{embedding}; however, the methods remain heuristic and solve each
request in isolation.

\subsection{Attack-Aware and Dynamically Defensive Placement}

A third group of works admits the idea that the security state of the
infrastructure is itself a dynamic variable. Iffl\"ander, Beierlieb,
and Kounev present an \textit{attack-aware} security function chain
reordering framework that changes the order of security functions
(e.g., firewall, IDS, DPI) as a function of observed malicious
traffic, empirically showing that chain order materially affects
detection performance and that no single static order dominates in
every situation~\cite{ifflander_attack-aware_2024}. Jia~et~al.
combine federated learning with SFC in mobile edge computing in the
FL-SFC framework, using a CNN-GRU anomaly-detection model at the edge
and a sparsification-plus-quantization compression scheme to lower
communication cost during training; the resulting pipeline produces
dynamic service policies and detects
anomalies~\cite{jia_efficient_2025}. Song~et~al. go further by coupling
a consortium blockchain for malicious-device detection with a DRL
placement agent (CB-DRL) built around an LSTM sequence-to-sequence
policy, demonstrating improvements in load-balancing and acceptance
ratio under DDoS-attack traffic in an IoT
scenario~\cite{song_end--end_2025}.

These papers are the closest in spirit to the present thesis: they
acknowledge that placement must react to the security state of the
infrastructure.

A key limitation, however, cuts across the security-aware works reviewed
in this section: every method is heuristic, meta-heuristic, or
optimization-based. These approaches solve each incoming SFC in
isolation and cannot learn from the arriving request stream, so they
adapt poorly to dynamic traffic patterns and shifting load
distributions. A learning-based agent --- specifically a DRL agent
that observes security information as part of its state --- is
therefore needed to exploit security awareness
\emph{adaptively} over time.

\section{Deep Reinforcement Learning for SFC Placement}
\label{sec:lr_drl}

The fourth component of the thesis title --- \textit{Deep Reinforcement
Learning} --- rests on the observation that DRL is now the dominant
learning-based approach to online SFC placement. The volume, diversity,
and recency of the works surveyed below (all of them from 2023--2025)
validate the choice of DRL as the methodological backbone of the
thesis. Equally important, however, is the observation that \emph{no
single DRL placement framework in this body of work is simultaneously
hard-constrained, topology-aware via graph neural networks, and
driven by a dynamic security-posture reward} --- which is the combination
that the thesis will introduce in Chapter~4. The reviewed works differ
in their choice of algorithm, their treatment of the action space, and
how much additional structure (parallelism, instance sharing,
attention) they bring to the policy, but they converge on a common set
of gaps that will be made explicit in
Section~\ref{sec:lr_synthesis}.

\subsection{Value-Based and Dueling Approaches}

Ran, Wang, and Hu propose MQDR, a dynamic deployment and
re-adjustment method built on a deep Q-network coupled with an
M-shortest-path algorithm. Two cooperating agents --- one for
deployment, one for readjustment --- share the substrate graph and
use MSPA to prune the action space; the authors report a
${\approx}25\%$ acceptance-ratio improvement over vanilla DQN and
a ${\approx}9\%$ improvement over load-balancing shortest-path
routing~\cite{ran_dynamic_2023}. Zhang~et~al. address a fundamental
weakness of one-step DQN for SFC placement by introducing DQN-MSRA, a
\emph{multistep} reinforcement learning variant that aggregates
rewards over a window, explicitly designed to emphasize long-horizon
outcomes over immediate-step rewards; a univariate modelling of
placement and routing further reduces the action space and avoids
exploration of invalid paths~\cite{zhang_dqn-msra_2025}. Taktak,
Escheikh, and Barkaoui tailor DQN to user-centric QoE/QoS-driven
SFC embedding for video-streaming traffic, introducing a QoE
threshold on the running 100-episode mean as a quality-of-learning
criterion and reporting placements that approach the optimum found
by exhaustive search~\cite{taktak_drl_2024}.

These value-based methods are effective but suffer from two common
limitations: they enforce constraints through penalty terms rather
than action masking, so infeasible actions remain reachable during
both training and inference; and they do not model security as an
explicit state dimension.

\subsection{Policy-Gradient, Actor-Critic, and Hybrid Approaches}

Xu~et~al. apply the asynchronous advantage actor-critic (A3C)
algorithm to MEC SFC placement with two additional mechanisms:
reuse of already-initialized VNFs and parallelization of the SFC into
sub-flows processed by duplicate VNF instances, reporting a
6--24\% improvement in quality-of-service over state-of-the-art
clustering methods~\cite{xu_dynamic_2023}. Zahedi, Mollahoseini, and
Heidary-Sharifabad's DSPVR system targets fog-cloud deployments and
solves the SFC placement problem through a DRL agent that prioritizes
the reuse of existing VNF instances, consistently beating DRL-SFCP
and DDQP on monetary cost~\cite{zahedi_dspvr_2023}. Khemili~et~al.
propose PBDRL, a \emph{parallel bi-state} DRL model that couples a
Markov Decision Process module for state-transition dynamics with an
LSTM that captures the long-term chain history, achieving
substantially higher cumulative reward than a ``VNF deep'' baseline
and a first-fit heuristic~\cite{khemili_parallel_2024}.

Policy-gradient methods such as these are better suited to
high-dimensional discrete action spaces than plain DQN. However, they
still do not enforce hard constraints by construction, none of them
model security as a dynamic placement concern, and all require
careful reward engineering to balance competing objectives. The
Proximal Policy Optimization (PPO) family that this thesis builds on
addresses the first of these issues by allowing direct integration of
action masks into the policy.

\subsection{DRL with Attention Mechanisms and Parallelism}

As SFCs grow more complex, placement policies have increasingly
adopted attention mechanisms and learned parallelism. He~et~al.
leverage an \emph{attention mechanism} within a DRL policy to jointly
solve VNF placement and routing, attending dynamically to the most
relevant substrate-graph regions at each decision
step~\cite{he_leveraging_2023}. Wang's fairness-aware throughput
maximization strategy applies DRL to parallelized VNFs in MEC, with
the policy trained to both parallelize and place chains under
fluctuating resource demand, improving average latency by $4.3\%$ and
deployment rate by $1.98\%$ over the best published
baseline~\cite{wang_fairness-aware_2025}. Lin~et~al.'s LPPIF
framework transforms a serial SFC into a parallelized SFC by
identifying VNF dependencies and then solves placement using DRL,
with a flexible resource-allocation mechanism and reuse of VNF
instances that jointly improve service latency and the acceptance
ratio of incoming SFC requests~\cite{lin_lppif_2025}.

This family of works establishes the value of \emph{topology-aware}
and \emph{structure-aware} DRL policies for SFC placement, and
motivates the use of graph neural networks --- the next natural step
--- as the feature extractor of the DRL policy. The thesis presented
here adopts exactly this direction while additionally enforcing hard
constraints through action masking and incorporating a security
posture reward.

\section{Classical Optimization and Heuristic Approaches}
\label{sec:lr_classical}

The third component of the thesis title --- \textit{Intelligent} ---
is intended in a precise sense: it designates a learning-based
policy that improves with experience and generalizes across request
distributions, as opposed to the classical algorithmic approaches
reviewed in this section. The reviewed classical works are both strong
baselines and, as surveyed below, a source of the formulations on
which DRL methods rely; however, none of them learn from the request
stream or adapt their parameters to an evolving substrate state ---
which is the deficit the word ``intelligent'' in the title is meant to
fill.

\subsection{ILP and Learning-Augmented Optimization}

Vishesh~et~al. combine an LSTM traffic-pattern predictor with an
Integer Linear Program for SFC placement on a real KubeEdge cluster,
using the LSTM's forecasts to feed the ILP with realistic future
resource demands. The resulting LSTM-ILP scheduler reports a $12.5\%$
improvement in response time over the default Kubernetes scheduler at
a workload of 50{,}000 requests~\cite{vishesh_optimized_2025}.
Lin~et~al. study the \emph{energy-aware} SFC embedding problem under
hierarchical resource allocation, formulating an ILP and three
polynomial-time heuristics that exploit device sleep states to reduce
energy while preserving acceptance
ratio~\cite{lin_energy-aware_2023}. Saidi~et~al. prove the VNF
placement-and-chaining problem NP-complete for both fixed-size
infrastructures and constant-size SFCs, and introduce a
constrained-shortest-path heuristic (CSPH-PG) over a
resource-homogenizing multi-partite graph that converges rapidly
toward the ILP optimum~\cite{saidi_constrained_2024}.

These works illustrate two distinct uses of classical optimization:
as an \textit{offline oracle} that establishes achievable bounds and
as an \textit{online heuristic} carefully tuned for a particular
sub-problem. Their principal limitation for our purposes is the same
as in the introduction: they do not learn from the request stream and
they do not treat security as a dynamic state variable.

\subsection{Meta-Heuristics and Auctions}

When exact optimization is intractable and no training data is
available, meta-heuristics remain a pragmatic choice.
Tam~et~al. develop a subswarm-guided ant-colony optimization (SgACO)
for multi-objective VNF placement and routing with an enhanced
pheromone-update rule and a beam-search refinement, showing that the
method satisfies many requests simultaneously and outperforms
competitors on COGENT, CONUS, and NSF
topologies~\cite{tam_subswarm-guided_2024}. Macedo~et~al. take an
economic view: they propose a distributed \emph{auction}-based SFC
placement algorithm for multi-domain 5G networks, allowing domains
to outsource service provisioning to peer domains and reporting
${\approx}20\%$ service-provider profit
gains~\cite{macedo_distributed_2024}.

Meta-heuristics and auctions are highly flexible but sacrifice the
millisecond-level responsiveness that an online DRL agent enjoys
once trained.

\subsection{Specialized Objective Heuristics}

A final cluster of works targets narrow but operationally important
objectives. Han, Meng, and Fan analyze the cloud-edge placement
setting and show that it is preferable to place an entire SFC
\textit{either} fully in the cloud or fully at the edge; they then
introduce a VNF performance-resource (P-R) function to describe
how computing performance scales with allocated resources, and
propose the RMET heuristic to maximize edge
throughput~\cite{han_sfc_2024}. Liang~et~al. examine the often
overlooked effect of \emph{packet-loss position}: their Palos
scheduler accounts for the fact that packet drops late in the SFC
waste all upstream processing, reducing packet-dropping cost by up to
$42.73\%$ and average latency by
$33.03\%$~\cite{liang_position-aware_2024}. Mohamad and Hassanein
target time-critical SFCs at the edge with the PPPS scheme, which
prioritizes premium services over best-effort ones and, via a
prediction-based look-ahead window, \emph{preempts} resources held by
best-effort services when necessary, driving the premium-service
rejection rate to approximately
zero~\cite{mohamad_preemptive_2024}.

These specialized heuristics identify dimensions of the placement
problem --- energy, packet loss, time-critical prioritization --- that
a learning-based framework should ideally be able to integrate into a
unified objective. This thesis demonstrates one such integration for
the security dimension.

\section{Conclusion}
\label{sec:lr_synthesis}

\begin{table}[!ht]
\centering
\caption{Comparative summary of recent SFC-placement
literature. ``Hard Constr.'' indicates guaranteed
feasibility via mechanisms such as action masking; ``Graph Policy''
indicates whether the policy explicitly exploits the substrate graph
structure; ``Fair Eval.'' indicates whether comparison aligns the
request stream across algorithms. (\checkmark: yes; \textendash: no;
$\circ$: partial.)}
\label{tab:lr_comparison}
\footnotesize
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.15}
\newcommand{\rothead}[1]{\rotatebox[origin=c]{60}{\textbf{#1}}}
\begin{tabularx}{\textwidth}{@{}p{3.1cm} X *{4}{c}@{}}
\toprule
\textbf{Work} & \textbf{Technique} & \rothead{Sec.} & \rothead{Hard Constr.} & \rothead{Graph Policy} & \rothead{Fair Eval.} \\
\midrule
Yang \textit{et al.}~\cite{yang_efficient_2024}            & Heuristic (ESSFCD-DO)                    & \checkmark   & \textendash & \textendash & \textendash \\
Sun \textit{et al.}~\cite{sun_security-aware_2023}         & Security-time scheduling                 & \checkmark   & \textendash & \textendash & \textendash \\
Dubba \& Killi~\cite{dubba_security-aware_2024}            & PSO / GLPSO                              & \checkmark   & \textendash & \textendash & \textendash \\
Iffl\"ander \textit{et al.}~\cite{ifflander_attack-aware_2024} & Attack-aware reordering                & \checkmark   & \textendash & \textendash & \textendash \\
Wang \textit{et al.}~\cite{wang_security-aware_2024}       & APPF-DSI (dedicated protection)          & \checkmark   & \textendash & \textendash & \textendash \\
Jia \textit{et al.}~\cite{jia_efficient_2025}              & FL + CNN-GRU                             & \checkmark   & \textendash & \textendash & \textendash \\
Song \textit{et al.}~\cite{song_end--end_2025}             & Blockchain + DRL (LSTM)                  & \checkmark   & \textendash & $\circ$     & \textendash \\
Wang \textit{et al.}~\cite{wang_towards_2025}              & SCB-SD (log-approx.)                     & \checkmark   & \textendash & \textendash & \textendash \\
\midrule
Ran \textit{et al.}~\cite{ran_dynamic_2023}                & DQN + M-shortest path                    & \textendash  & \textendash & \textendash & \textendash \\
He \textit{et al.}~\cite{he_leveraging_2023}               & DRL + attention                          & \textendash  & \textendash & $\circ$     & \textendash \\
Zhang \textit{et al.}~\cite{zhang_dqn-msra_2025}           & DQN + multi-step RL                      & \textendash  & \textendash & \textendash & \textendash \\
Taktak \textit{et al.}~\cite{taktak_drl_2024}              & QoE/QoS DQN                              & \textendash  & \textendash & \textendash & \textendash \\
Xu \textit{et al.}~\cite{xu_dynamic_2023}                  & A3C + VNF reuse + parallelism            & \textendash  & \textendash & \textendash & \textendash \\
Zahedi \textit{et al.}~\cite{zahedi_dspvr_2023}            & DSPVR (DRL + reuse)                      & \textendash  & \textendash & \textendash & \textendash \\
Khemili \textit{et al.}~\cite{khemili_parallel_2024}       & Bi-state (MDP + LSTM)                    & \textendash  & \textendash & \textendash & \textendash \\
Wang~\cite{wang_fairness-aware_2025}                        & DRL + NF parallelism                     & \textendash  & \textendash & $\circ$     & \textendash \\
Lin \textit{et al.}~\cite{lin_lppif_2025}                  & DRL + VNF share + parallelism            & \textendash  & \textendash & $\circ$     & \textendash \\
\midrule
Vishesh \textit{et al.}~\cite{vishesh_optimized_2025}      & LSTM + ILP                               & \textendash  & \checkmark  & \textendash & \textendash \\
Macedo \textit{et al.}~\cite{macedo_distributed_2024}      & Distributed auction                      & \textendash  & \textendash & \textendash & \textendash \\
Han \textit{et al.}~\cite{han_sfc_2024}                    & VNF P-R function + RMET                  & \textendash  & \textendash & \textendash & \textendash \\
Tam \textit{et al.}~\cite{tam_subswarm-guided_2024}        & Ant-colony (SgACO)                       & \textendash  & \textendash & \textendash & \textendash \\
Saidi \textit{et al.}~\cite{saidi_constrained_2024}        & CSPH-PG (constrained SP)                 & \textendash  & \checkmark  & \textendash & \textendash \\
Lin \textit{et al.}~\cite{lin_energy-aware_2023}           & Energy-aware ILP + heuristics            & \textendash  & \checkmark  & \textendash & \textendash \\
Liang \textit{et al.}~\cite{liang_position-aware_2024}     & Palos (pos.-aware)                       & \textendash  & \textendash & \textendash & \textendash \\
Mohamad \& Hassanein~\cite{mohamad_preemptive_2024}        & PPPS (preemptive pred.)                  & \textendash  & \textendash & \textendash & \textendash \\
\midrule
\textbf{This thesis} & \textbf{Maskable PPO + GNN + CIA triad + zones + link sec.} & \textbf{\checkmark} & \textbf{\checkmark} & \textbf{\checkmark} & \textbf{\checkmark} \\
\bottomrule
\end{tabularx}
\end{table}

Two complementary observations emerge from
Table~\ref{tab:lr_comparison}. \textbf{First}, every work that
incorporates security into the placement objective --- Yang~et~al.,
Sun~et~al., Dubba and Killi, both Wang groups, Iffl\"ander~et~al.,
Jia~et~al., and
Song~et~al.~\cite{yang_efficient_2024,sun_security-aware_2023,dubba_security-aware_2024,wang_security-aware_2024,wang_towards_2025,ifflander_attack-aware_2024,jia_efficient_2025,song_end--end_2025}
--- relies on heuristic, meta-heuristic, or optimization-based methods
(ILP, particle-swarm search, augmenting-path algorithms,
chain-reordering rules, federated-learning anomaly detectors). None of
these approaches learns from the arriving request stream; each
incoming SFC is solved in isolation, so the placement policy cannot
adapt to changing traffic patterns or shifting load distributions.
\textbf{Second}, every DRL-based
placement
agent~\cite{ran_dynamic_2023,he_leveraging_2023,zhang_dqn-msra_2025,taktak_drl_2024,xu_dynamic_2023,zahedi_dspvr_2023,khemili_parallel_2024,wang_fairness-aware_2025,lin_lppif_2025}
ignores security requirements entirely: the \emph{Sec.} column
carries no \checkmark\ for any of these works, and security plays no
role in their state representation, reward signal, or constraint
enforcement. The intersection of the two research
directions --- \emph{security-aware} and \emph{DRL-based} --- is
therefore empty, leaving a clear gap for Deep Reinforcement Learning
applied to security-posture-aware SFC placement.

Beyond this primary gap, the table reveals additional deficiencies
that compound the problem. Among
the DRL-based approaches, constraints are consistently enforced
through penalty rewards rather than hard action masking, graph
structure is exploited only through partial mechanisms such as
attention weights or parallelism-aware encoders, and none of the
reviewed works align the random-number stream across algorithms to
isolate algorithmic gain from evaluation variance.

This thesis closes the primary gap by applying DRL to
security-posture-aware SFC placement, and simultaneously addresses
the secondary deficiencies: Maskable PPO with action masking
guarantees hard-constraint feasibility across CIA, zone, and
link-security constraints; a GNN feature extractor provides principled
topology awareness; a security-aware reward signal incentivises
placements with CIA posture exceeding the hard minimum; and an
aligned-RNG evaluation protocol ensures fair comparison --- as the
final row of Table~\ref{tab:lr_comparison} summarises.
