\begin{thebibliography}{999}
\addcontentsline{toc}{chapter}{References}

\bibitem{etsi2013}
ETSI Industry Specification Group (ISG) NFV, ``Network Functions Virtualisation: An Introduction, Benefits, Enablers, Challenges and Call for Action,'' \textit{White Paper}, Oct. 2013.

\bibitem{mckeown2008}
N. McKeown, T. Anderson, H. Balakrishnan, G. Parulkar, L. Peterson, J. Rexford, S. Shenker, and J. Turner, ``OpenFlow: Enabling innovation in campus networks,'' \textit{ACM SIGCOMM Computer Communication Review}, vol. 38, no. 2, pp. 69--74, 2008.

\bibitem{mijumbi2016}
R. Mijumbi, J. Serrat, J.-L. Gorricho, N. Bouten, F. De Turck, and R. Boutaba, ``Network function virtualization: State-of-the-art and research challenges,'' \textit{IEEE Communications Surveys \& Tutorials}, vol. 18, no. 1, pp. 236--262, 2016.

\bibitem{halpern2015}
J. Halpern and C. Pignataro, ``Service Function Chaining (SFC) Architecture,'' IETF RFC 7665, Oct. 2015.

\bibitem{nfv5g2017}
R. Mijumbi, J. Serrat, J.-L. Gorricho, J. Rubio-Loyola, and S. Davy, ``Server placement and assignment in virtualized radio access networks,'' in \textit{Proc. 10th IFIP/IEEE International Symposium on Integrated Network Management}, 2017.

\bibitem{dolev2013}
S. Dolev, S. Frenkel, M. Kumar, and P. Shor, ``Placing Virtual Network Functions: Computational Complexity and Algorithms,'' \textit{IEEE Transactions on Network and Service Management}, 2013.

\bibitem{yu2008}
M. Yu, Y. Yi, J. Rexford, and M. Chiang, ``Rethinking virtual network embedding: Substrate support for path splitting and migration,'' \textit{ACM SIGCOMM Computer Communication Review}, vol. 38, no. 2, pp. 17--29, 2008.

\bibitem{cohen2015}
R. Cohen, L. Lewin-Eytan, J. S. Naor, and D. Raz, ``Near optimal placement of virtual network functions,'' in \textit{Proc. IEEE INFOCOM}, 2015, pp. 1346--1354.

\bibitem{bari2015}
M. F. Bari, S. R. Chowdhury, R. Ahmed, and R. Boutaba, ``On orchestrating virtual network functions,'' in \textit{Proc. 11th International Conference on Network and Service Management (CNSM)}, 2015, pp. 50--56.

\bibitem{bhamare2016}
D. Bhamare, R. Jain, M. Samaka, and A. Erbad, ``A survey on service function chaining,'' \textit{Journal of Network and Computer Applications}, vol. 75, pp. 138--155, 2016.

\bibitem{rankothge2017}
W. Rankothge, J. Ma, F. Le, A. Russo, and J. Lobo, ``Optimizing resource utilization in network function virtualization,'' in \textit{Proc. IEEE NOMS}, 2017.

\bibitem{mao2016}
H. Mao, M. Alizadeh, I. Menache, and S. Kandula, ``Resource management with deep reinforcement learning,'' in \textit{Proc. 15th ACM Workshop on Hot Topics in Networks}, 2016, pp. 50--56.

\bibitem{yan2020}
Z. Yan, J. Ge, Y. Wu, L. Li, and T. Li, ``Automatic virtual network embedding: A deep reinforcement learning approach with graph convolutional networks,'' \textit{IEEE Journal on Selected Areas in Communications}, vol. 38, no. 6, pp. 1040--1057, 2020.

\bibitem{heo2020}
D. Heo, S. Cho, B. Han, and H. Yoon, ``Service function chain embedding with deep Q-networks,'' in \textit{Proc. IEEE International Conference on Communications (ICC)}, 2020.

\bibitem{dolati2021}
M. Dolati, S. B. Hassanpour, M. Gharbani, and A. Khonsari, ``DeepViNE: Virtual network embedding with deep reinforcement learning,'' in \textit{Proc. IEEE INFOCOM WKSHPS}, 2021.

\bibitem{schulman2017}
J. Schulman, F. Wolski, P. Dhariwal, A. Radford, and O. Klimov, ``Proximal policy optimization algorithms,'' \textit{arXiv preprint arXiv:1707.06347}, 2017.

\bibitem{schulman2016}
J. Schulman, P. Moritz, S. Levine, M. Jordan, and P. Abbeel, ``High-dimensional continuous control using generalized advantage estimation,'' in \textit{Proc. International Conference on Learning Representations (ICLR)}, 2016.

\bibitem{raffin2021}
A. Raffin, A. Hill, A. Gleave, A. Kanervisto, M. Ernestus, and N. Dormann, ``Stable-baselines3: Reliable reinforcement learning implementations,'' \textit{Journal of Machine Learning Research}, vol. 22, no. 268, pp. 1--8, 2021.

\bibitem{sb3contrib2021}
A. Kanervisto, C. Scheller, and V. Hautam\"{a}ki, ``Action masking in discrete action spaces in deep reinforcement learning,'' in \textit{Proc. IEEE Conference on Games (CoG)}, 2020.

\bibitem{huang2020}
S. Huang and S. Ontañón, ``A closer look at invalid action masking in policy gradient algorithms,'' \textit{arXiv preprint arXiv:2006.14171}, 2020.

\bibitem{vinyals2019}
O. Vinyals, I. Babuschkin, W. M. Czarnecki, M. Mathieu, A. Dudzik, J. Chung, D. H. Choi, R. Powell, T. Ewalds, P. Georgiev, \textit{et al.}, ``Grandmaster level in StarCraft II using multi-agent reinforcement learning,'' \textit{Nature}, vol. 575, pp. 350--354, 2019.

\bibitem{kipf2017}
T. N. Kipf and M. Welling, ``Semi-supervised classification with graph convolutional networks,'' in \textit{Proc. International Conference on Learning Representations (ICLR)}, 2017.

\bibitem{velickovic2018}
P. Veli\v{c}kovi\'{c}, G. Cucurull, A. Casanova, A. Romero, P. Li\`{o}, and Y. Bengio, ``Graph attention networks,'' in \textit{Proc. International Conference on Learning Representations (ICLR)}, 2018.

\bibitem{hamilton2017}
W. Hamilton, Z. Ying, and J. Leskovec, ``Inductive representation learning on large graphs,'' in \textit{Proc. Advances in Neural Information Processing Systems (NeurIPS)}, 2017, pp. 1024--1034.

\bibitem{vaswani2017}
A. Vaswani, N. Shazeer, N. Parmar, J. Uszkoreit, L. Jones, A. N. Gomez, \L{}. Kaiser, and I. Polosukhin, ``Attention is all you need,'' in \textit{Proc. Advances in Neural Information Processing Systems (NeurIPS)}, 2017, pp. 5998--6008.

\bibitem{lotfi2022}
M. Lotfi and P. Sioshansi, ``Graph neural network approaches for virtual network embedding,'' \textit{IEEE Transactions on Network and Service Management}, 2022.

\bibitem{shi2021}
Y. Shi, K. Letaief, R. Schober, and Z. Han, ``Graphically-aware resource allocation for wireless networks,'' \textit{IEEE Transactions on Wireless Communications}, 2021.

\bibitem{rusek2020}
K. Rusek, J. Su\'{a}rez-Varela, A. Mestres, P. Barlet-Ros, and A. Cabellos-Aparicio, ``Unveiling the potential of graph neural networks for network modeling and optimization in SDN,'' in \textit{Proc. ACM SOSR}, 2020.

\bibitem{almasan2019}
P. Almasan, J. Su\'{a}rez-Varela, A. Badia-Sampera, K. Rusek, P. Barlet-Ros, and A. Cabellos-Aparicio, ``Deep reinforcement learning meets graph neural networks: exploring a routing optimization use case,'' \textit{arXiv preprint arXiv:1910.07421}, 2019.

\bibitem{guo2021}
J. Guo, C. Jiang, Y. Qian, and P. Wan, ``Joint resource allocation and task scheduling for heterogeneous Internet of Things using graph neural network,'' \textit{IEEE Internet of Things Journal}, 2021.

\bibitem{pattaranantakul2016}
M. Pattaranantakul, R. He, A. Meddahi, and Z. Zhang, ``SecMANO: Towards network functions virtualization (NFV) based security MANagement and Orchestration,'' in \textit{Proc. IEEE TrustCom/BigDataSE/ISPA}, 2016.

\bibitem{yu2021}
Y. Yu, J. Li, and S. Guo, ``Security-aware service function chain embedding in NFV-enabled networks,'' \textit{IEEE Transactions on Network and Service Management}, vol. 18, no. 2, 2021.

\bibitem{barlow1975}
R. E. Barlow and F. Proschan, \textit{Statistical Theory of Reliability and Life Testing}. New York: Holt, Rinehart and Winston, 1975.

\bibitem{yago2019}
M. Yago, J. M. Alcaraz, and M. V. Bueno-Delgado, ``Stochastic approaches to network reliability evaluation,'' \textit{Reliability Engineering \& System Safety}, vol. 188, pp. 450--460, 2019.

\bibitem{zhu2021}
Z. Zhu, M. Huang, M. Ye, J. Fan, and J. Wu, ``DDPG-based virtual network function migration in network function virtualization,'' in \textit{Proc. IEEE ICCC}, 2021.

\bibitem{viterbi1967}
A. J. Viterbi, ``Error bounds for convolutional codes and an asymptotically optimum decoding algorithm,'' \textit{IEEE Transactions on Information Theory}, vol. 13, no. 2, pp. 260--269, 1967.

\end{thebibliography}
