\newpage
\begin{center}
\LARGE\textbf{{Acknowledgements}}\\
\end{center}
\vspace{0.3in}
\begin{center}
\textbf{Praise be to ALLAH, the Cherisher and Lord\\of the World, Most Gracious
and Most Merciful}
\end{center}
\normalsize{First and foremost, I would like to thank ALLAH Almighty (the Most
Beneficent and Most Merciful) for granting me the strength, knowledge, ability,
and opportunity to undertake this research and to persevere and complete it
satisfactorily. Without His countless blessings, this achievement would not have
been possible. May His peace and blessings be upon His messenger Hazrat Muhammad
(PBUH), upon his family, companions, and all who follow him.

I wish to express my deepest and most sincere gratitude to my supervisor, Dr.
Muhammad Hasanain Chaudary, for his invaluable guidance, continuous
encouragement, and insightful feedback throughout the course of this research.
His expertise in computer networks and machine learning has been instrumental in
shaping this work. I am deeply indebted to him for the time and effort he
invested in my academic and professional development.

I would also like to acknowledge the Department of Computer Science at COMSATS
University Islamabad for providing an excellent academic environment and
computational resources that made this research possible.

My heartfelt gratitude goes to my parents, whose unconditional love, prayers,
and sacrifices have been a constant source of motivation. Their unwavering
belief in me has been my greatest strength throughout this journey.

I am also thankful to my colleagues and friends who provided moral support,
constructive discussions, and encouragement throughout my studies.

Finally, I acknowledge the open-source communities behind Stable-Baselines3,
PyTorch Geometric, Gymnasium, and NetworkX, whose tools formed the backbone of
the software infrastructure of this research.}

\vspace{0.5in}
\hspace{265pt}Ali Raheel

\hspace{240pt}CIIT/SP24-RCS-001/LHR
